Wealth Inequality in China: Evidence from the 2017 and 2019 CHFS

### 2025-04-25 report/draft.qmd
update

# Introduction
China's remarkable economic growth over the past five decades was initiated by the reform and opening-up policy launched in 1978, transitioning the country from a centrally planned to a market-oriented economy. From the openning up in 1978 to 2024, China's GDP growth rate averaged over 9% per year, and 800 million people were lifted out of poverty, with considerable improvment in the nation's health and education [@world_bank_world_2024]. This growth, however, has not been uniform across time, geography, or economic sectors. Coastal provinces such as Jiangsu and Guangdong experienced substantially higher growth rates, benefiting from early exposure to foreign investment and trade, while inland and rural areas lagged, leading to pronounced regional disparities [@liao_sixty_2016]. Sectorally, China's strong growth has been driven by fixed-capital investment and export-oriented manufacuring, which althoug effective in the past, start to show signs of both domestic and international imbalances [@world_bank_world_2024]. Since 2010, China's annual GDP growth rate has been declining from 10.6% in 2011 to 5.2% in 2023[@world_bank_gdp_2025].

Concurrently, China experienced a substantial rise in wealth and income inequality. The wealth share controlled by the top 10% increased markedly from approximately 40% in the late 1970s to nearly 70% by 2015, while the share held by the bottom half significantly decreased [@piketty_capital_2019]. Some studies indicate that China's overall inequality reached its peak around 2010, after which it started to gradually decline [@kanbur_great_2021; @zhang_survey_2021]. Additionally, it is well-established that China's inequality closely correlates with its urban-rural and coastal-inland divisions, with urban and coastal regions exhibiting substantially higher levels of income and wealth compared to their rural and inland counterparts [@piketty_capital_2019; @zhang_survey_2021].

This paper analyzes the contemporary landscape of wealth inequality in China using recent data from the 2017 and 2019 China Household Finance Survey (CHFS). It specifically investigates whether recent trends confirm claims of continued inequality reduction since 2010. Through factor decomposition, the study examines the sources of wealth inequality within the entire population as well as separately within urban-rural and coastal-inland subgroups, highlighting variations in inequality among these different groups. Furthermore, the study applies group decomposition methods to assess the extent to which rural-urban and coastal-inland disparities contribute to overall wealth inequality. Through these detailed decompositions and analyses, the paper aims to provide a comprehensive and nuanced understanding of current patterns and dynamics in China's wealth inequality.

# Background
This section presents some notable findings on China's income and wealth inequality in the existing literature.

China has experienced a considerable increase in income and wealth inequality since the reform and opening-up policy initiated in 1978. Using diverse sources, including tax records, household surveys, and private wealth rankings, @piketty_capital_2019 provide a detailed analysis of the transformation in China's inequality profile. Their findings show that between 1978 and 2015, the income share of the top 10% increased from 27% to 41%, while the share of the bottom 50% decreased from 27% to 15%. Wealth concentration also intensified significantly: the wealth share held by the top 10% rose from around 40% to nearly 70%, whereas the bottom 50%'s share dropped from over 15% to about 5%. Additionally, they documented an increase in the urban-to-rural per capita income ratio, from less than 200% in 1978 to approximately 350% by 2015. Humorously noting this stark shift, they remarked that China had transitioned from Nordic-level equality to U.S.-level inequality during the reform era.

@knight_why_2022 further document a rapid rise in China's household wealth inequality between 2002 and 2013, reporting that the national Gini coefficient for household wealth per capita increased markedly from 0.50 to 0.62. They identify housing wealth as the primary driver behind this increase, with its share rising from 53% to 73% of total household wealth, and its contribution to overall wealth inequality growing from 64% to 79%. The authors emphasize that differential savings rates across income groups, uneven house price inflation—particularly pronounced in major urban centers—and widening urban-rural disparities significantly intensified wealth inequality during this period.

The income and wealth inequality highly intersects with urban-rural and reginal disparities. @zhang_survey_2021 highlights significant rural-urban and regional disparities underlying China's rising inequality since its economic reforms. The urban-rural income ratio increased sharply, peaking at around 3.3 in 2009 before declining slightly to 2.7 by 2019, still substantially higher than pre-reform levels. Despite a notable decline in overall inequality after 2010—with the national Gini coefficient decreasing from a peak of 0.491 in 2008 to approximately 0.465 in 2019—regional inequality remains pronounced. For instance, per capita GDP in the eastern coastal region was roughly 2.6 times higher than in the western region by 2018. Additionally, rural-to-urban migration grew dramatically from fewer than 20 million in 1990 to approximately 290 million by 2019, reflecting both the economic disparity between regions and the persistent urban-rural divide, exacerbated by restricted access to urban public services under the Hukou household registration system.

Consistent with the finding in @zhang_survey_2021 that China's inequality peaked in 2008 and 2009, @kanbur_great_2021 similarly document that China's inequality increased sharply after economic reforms but started to plateau around 2010. They find the national Gini coefficient peaked at about 0.525 in 2010, subsequently declining to approximately 0.476 in 2016 before rising slightly again. Consistent with @zhang_survey_2021, they attribute this recent turnaround primarily to narrowing rural-urban and coastal-inland disparities, driven by demographic changes, tightening rural labor markets, and rising rural wages.

# Data
The analysis in this study utilizes data from the 2017 and 2019 waves of the China Household Finance Survey (CHFS), a nationally representative dataset collected by the Survey and Research Center for China Household Finance at Southwestern University of Finance and Economics [@gan_report_2014]. The CHFS dataset includes household-level financial information on income, assets, debts, consumption, and demographic characteristics. Data are collected through face-to-face interviews using a multi-stage probability proportional to size (PPS) sampling method, ensuring representativeness at both national and provincial levels. Sample weights provided by CHFS are calculated to account for the sampling design, adjusting for unequal probabilities of selection at different stages.

Several limitations of the CHFS dataset should be noted. First, it relies on self-reported values for income and assets, potentially introducing biases such as underreporting, recall inaccuracies, or privacy concerns [@gan_report_2014]. Second, definitional changes across survey waves may affect comparability. For instance, the 2019 survey separately records garage assets, previously included within housing assets in the 2017 wave, potentially complicating longitudinal analyses. Additionally, certain variables in the dataset are top-coded to protect respondent privacy, with extreme values replaced by preset caps, accompanied by censoring indicators [@chfs_2019_2019]. Such censoring is not accounted for in this study and might underestimate the level of income and wealth inequality in China.

The CHFS dataset has been extensively utilized in academic research. Studies leveraging the CHFS data have explored topics ranging from wealth distribution and inequality [@tan_earnings_2017], household finance in rural China [@gan_report_2014], household medical insurance expenditure [@liu_health_2022], and many others.

>>>>

# Introduction

China's economic trajectory over the past five decades, initiated by the reform and opening-up policy launched in 1978, involved a transition from a centrally planned to a more market-oriented economy. This period yielded high average GDP growth rates alongside considerable improvements in poverty reduction, health, and education [@world_bank_world_2024; @ravallion_chinas_2009]. This growth, however, manifested unevenly across time, geography, and economic sectors. Coastal provinces, for instance, often experienced higher growth rates, benefiting from earlier integration with global markets, while inland and rural areas developed more slowly, contributing to regional disparities [@liao_sixty_2016]. Sectorally, growth was initially propelled by fixed-capital investment and export-oriented manufacturing, a model that later showed signs of creating imbalances [@world_bank_world_2024]. China's annual GDP growth rate moderated from 10.6% in 2010 to 5.2% by 2023, reflecting a slowdown from the high rates experienced earlier in the decade [@world_bank_gdp_2025].

Parallel to this economic expansion, China experienced a substantial increase in wealth and income inequality. Analysis combining diverse data sources reveals a marked shift in distribution between 1978 and 2015; the wealth share controlled by the top 10% increased from approximately 40% to nearly 70%, while the share held by the bottom 50% diminished considerably [@piketty_capital_2019]. Wealth inequality escalated rapidly, with the Gini coefficient for household wealth per capita rising markedly between 2002 and 2013, reaching high levels by international comparison [@knight_why_2022; @tan_earnings_2017]. Some research suggests that overall income inequality may have reached a peak around 2008-2010, potentially followed by a gradual decline or stabilization, linked partly to narrowing structural divides [@zhang_survey_2021; @kanbur_great_2021]. It is well-documented that China's inequality dynamics intersect closely with urban-rural and coastal-inland divisions; urban and coastal regions consistently exhibit higher average levels of income and wealth compared to their rural and inland counterparts [@piketty_capital_2019; @zhang_survey_2021]. The urban-rural income ratio remained high even in the late 2010s, despite some narrowing from its 2009 peak, and regional income differences persist [@zhang_survey_2021], findings consistent across multiple data sources [@xie_income_2014].

This study contributes to the understanding of China's wealth inequality by employing recent microdata from the 2017 and 2019 China Household Finance Survey (CHFS). The analysis focuses specifically on assessing whether wealth inequality trends during this later period substantiate claims of ongoing inequality reduction observed in earlier research focused primarily on income. The paper utilizes factor decomposition methods to pinpoint the contributions of various asset types to overall wealth inequality. A further level of analysis involves applying this factor decomposition separately to distinct population subgroups defined by urban-rural residence and coastal-inland location, allowing for an examination of potential heterogeneity in the drivers of inequality across these dimensions. Additionally, group decomposition techniques are used to quantify the portion of total wealth inequality attributable to the average wealth differences between these urban-rural and coastal-inland groups. This combination of recent data and multiple decomposition strategies aims to offer a detailed perspective on the patterns and structural underpinnings of wealth inequality in contemporary China.

# Literature Review
This section reviews notable findings from existing literature on the evolution of China's income and wealth inequality following the market-oriented reforms initiated in 1978.

China's economic trajectory since 1978 has been characterized by high rates of growth alongside a considerable increase in economic disparities. Utilizing diverse data including tax records, surveys, and wealth rankings, @piketty_capital_2019 meticulously document this transformation. Their analysis reveals a substantial redistribution of national income between 1978 and 2015: the top 10% saw their income share rise from 27% to 41%, while the bottom 50%'s share conversely fell from 27% to 15%. Wealth concentration followed a similar, potentially more pronounced, pattern. The wealth share of the top 10% increased from approximately 40% to nearly 70%, diminishing the bottom 50%'s share from over 15% to around 5%. Concurrently, the urban-rural income gap widened, with the average per capita income ratio increasing from less than 2:1 to about 3.5:1. This shift led @piketty_capital_2019 to observe that China moved from equality levels comparable to Nordic countries towards those observed in the U.S. during this period. Complementing this, @xie_income_2014 established, using multiple surveys from the mid-2000s onwards, that China's Gini coefficient had reached levels between 0.53 and 0.55, exceeding official figures and the US benchmark at the time.

Focusing on the period between 2002 and 2013, @knight_why_2022 provide further evidence on the escalation of household wealth inequality, reporting an increase in the Gini coefficient for wealth per capita from 0.50 to 0.62. A key finding is the role of housing wealth, which grew from 53% to 73% of total household assets and accounted for an increasing share of wealth inequality, rising from 64% to 79% of the total. The authors attribute the intensification of wealth inequality during these years to a combination of factors, including differential saving rates favoring the wealthy, and uneven house price inflation which benefited homeowners in major urban centers.

These trends in income and wealth inequality are deeply intertwined with persistent structural divides within China. Significant rural-urban and regional disparities remain central to understanding China's inequality landscape [@zhang_survey_2021; @xie_income_2014]. The urban-rural income ratio, despite some decline after peaking around 3.3 in 2009, remained elevated at approximately 2.7 in 2019 [@zhang_survey_2021]. This gap is reinforced by the *hukou* (household registration) system, which limits rural migrants' access to urban services and opportunities, even as migration increased substantially from 1990 to 2019. Regional inequality also persists, with coastal areas generally outpacing inland regions [@liao_sixty_2016]. While overall income inequality showed signs of stabilization or decline after reaching a peak around 2008-2010 [@zhang_survey_2021; @kanbur_great_2021], attributed partly to narrowing structural gaps driven by demographic shifts and tightening rural labor markets, geographical disparities endure.

The transition towards a market economy itself introduced multiple drivers of inequality [@benjamin_income_2008; @zhang_survey_2021]. In rural China, the rise was associated with unequal access to non-agricultural income sources, particularly family businesses, alongside periods of slow growth or falling prices in the agricultural sector [@benjamin_income_2008; @ravallion_chinas_2009]. In urban areas, contributing factors included the phasing out of subsidies and entitlements, increasing wage dispersion driven by market forces and enterprise reforms, higher returns to education (especially tertiary degrees), and unemployment resulting from State-Owned Enterprise (SOE) restructuring [@benjamin_income_2008]. Uneven regional development was further amplified by policies favoring coastal regions, as well as the differential impacts of globalization and marketization [@liao_sixty_2016]. It is worth noting, however, that considerable inequality also exists within specific localities, suggesting that factors beyond broad regional or urban-rural classifications are also influential [@benjamin_income_2008].

Furthermore, the pattern of economic growth influenced distributional outcomes. Research indicates that growth originating in the agricultural and rural sectors was more effective at reducing both poverty and overall inequality compared to growth in the industrial or service sectors [@ravallion_chinas_2009]. This suggests the importance of the initial economic structure and the specific mechanisms linking growth to income gains across population segments. Conversely, the rise in inequality itself appeared to impede poverty reduction; provinces that began the reform period with higher levels of inequality experienced slower subsequent progress in reducing poverty, partly because growth was less effective at lifting incomes at the lower end of the distribution in these areas [@ravallion_chinas_2009]. Specific policies, such as the initial agrarian reforms and adjustments to agricultural procurement prices, also demonstrably affected poverty and inequality levels. While income inequality may have plateaued or slightly decreased post-2010, linked to narrowing structural gaps and potentially policy interventions [@zhang_survey_2021; @liao_sixty_2016], wealth concentration, influenced by real estate appreciation, continued its upward trend well into the 2010s [@knight_why_2022; @piketty_capital_2019].

# Data
The empirical analysis in this paper utilizes microdata from the 2017 and 2019 waves of the China Household Finance Survey (CHFS). The CHFS, conducted biennially by the Survey and Research Center for China Household Finance at Southwestern University of Finance and Economics (SWUFE) since 2011, is designed to collect household financial information representative at both national and provincial levels [@gan_report_2014]. The survey employs a stratified, three-stage Probability Proportional to Size (PPS) random sampling methodology. This approach involves selecting counties, followed by village or neighborhood committees, and then households, incorporating oversampling of high-wealth households to better capture asset distribution.

Data are gathered through face-to-face interviews using Computer-Assisted Personal Interviewing (CAPI), collecting information at both household and individual levels. As with most survey data, CHFS figures are subject to potential limitations. Asset and income data rely on self-reporting, which can be affected by reporting error or bias. Respondent privacy is protected through methods including the top-coding of certain high-value income and asset variables; this may result in an underestimation of the upper tails of the respective distributions and potentially impact inequality metrics. Researchers should also note that minor adjustments in variable definitions between the 2017 and 2019 waves, such as the separate classification of garage assets in 2019, may require consideration for longitudinal comparisons. Missing data are handled through specific coding and imputation for certain variables.

The CHFS provides sample weights, calculated based on the multi-stage sampling design and adjusted using population benchmarks, to facilitate representative analysis. The dataset has been utilized in numerous academic studies exploring various facets of the Chinese economy and household behavior. Prior research using CHFS data has examined topics such as income and wealth distribution [@tan_earnings_2017], rural household finance [@gan_report_2014], household responses to health shocks and insurance availability [@liu_health_2022] and many others [@tan_earnings_2017; @gan_report_2014; @liu_health_2022].

The 2017 CHFS wave surveyed approximately 38,994 households, while the 2019 wave included 34,643 households across 343 county-level units in 29 provinces [@chfs_2017_2017; @chfs_2019_2019]. For the present analysis, observations identified as low quality in the 2017 data, along with those lacking essential variables for this study, were omitted; however, the final analytical sample retains the substantial majority of the original observations.
